Guarded Recursion in Agda via Sized Types

### 2019-01-29 code-oneclock-paper/latex/Introduction.tex
Rewriting
Some rewriting (Sections 1-4)

Dependent type theory is an expressive functional programming
language that underlies the deductive system of proof assistants such
as Coq, Agda and Lean.  It is a total language,
meaning that every program definable inside type theory is necessarily
terminating. This is an important requirement that ensures the
consistency of the type system.

Possibly non-terminating computations
and infinite structures such as non-wellfounded trees can be
represented in type theory by extending the type system with
coinductive types. Recursively defined elements of these types
are required to be productive \cite{Coquand93} in the sense that every
finite unfolding can be computed in a finite number of steps.
In Agda's encoding of coinductive types using ``musical notation''
\cite{DA10}, productivity is enforced via a strict obligation: in the
definition of a corecursive function, the recursive
call is required to appear directly under
the application of a constructor. A similar syntactic
check is employed in Coq.
This restriction
typically makes programming with coinductive types cumbersome,
which spawned the search for alternative techniques for ensuring the
well-definedness of corecursive definitions.

We focus on two of these techniques in which productivity is encoded
at the level of types: sized types and guarded recursion. A sized type
\cite{Hughes96} is a type annotated with the number of unfolding that
elements of this type can undergo. Sized types have been implemented
in Agda \cite{A-sized,AVW-normalization} and can be used in
combination with coinductive records to specify
coinductive types \cite{Copatterns}.  Guarded recursion \cite{Nakano}
is a different approach where the type system is enhanced with a
modality, called ``later'' and written \IC{▻}, encoding time delay in
types. The later modality comes with a general fixpoint combinator for
programming with productive recursive functions and allows the
specification of guarded recursive types, \eg the type of guarded
streams satisfying the isomorphism \F{Str} \Ar{A} \F{≃} \Ar{A} \F{×}
\IC{▻} (\F{Str} \F{A}). The tail of a guarded stream is available only
one time step ahead from now, as evidenced by the application of the
later modality to \F{Str} \F{A}. Coinductive types can be constructed
from guarded recursive types in type theories that additionally
support clock variables and universal quantification over them
\cite{atkey2013productive,BahrGM17}. Currently there is no
implementation of a calculus with guarded recursion.
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory with guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aformentioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant'' .  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. In this respect \GTT\ is
closer in spirit to Atkey and McBride's language. The design of \GTT\
also has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

@@

This paper only include the essential parts of our Agda
formalization. The full code is available at
\url{https://github.com/niccoloveltri/agda-clott}. The formalization
uses Agda 2.5.4.1 and Agda standard library 0.16. The paper is
extracted from a literate Agda file, which implies that all the
displayed code has passed Agda's type-check.

>>>>

Dependent type theory is an expressive functional programming
language that underlies the deductive system of proof assistants such
as Coq, Agda, and Lean \cite{norell}.  It is a total language,
meaning that every program definable inside type theory is necessarily
terminating. This is an important requirement that ensures the
consistency of the type system.

Possibly non-terminating computations
and infinite structures such as non-wellfounded trees can be
represented in type theory by extending the type system with
coinductive types. Recursively defined elements of these types
are required to be productive in the sense that every
finite unfolding can be computed in a finite number of steps \cite{Coquand93}.
In Agda's encoding of coinductive types using ``musical notation''
\cite{DA10}, productivity is enforced via a strict obligation: in the
definition of a corecursive function, the recursive
call is required to appear directly under
the application of a constructor. A similar syntactic
check is employed in Coq.
This restriction
typically makes programming with coinductive types cumbersome,
which spawned the search for alternative techniques for ensuring the
well-definedness of corecursive definitions.

We focus on two of these techniques in which productivity is encoded
at the level of types: sized types and guarded recursion. A sized type
 is a type annotated with the number of unfolding that
elements of this type can undergo \cite{Hughes96}. Sized types have been implemented
in Agda and can be used in
combination with coinductive records to specify
coinductive types \cite{Copatterns,AVW-normalization,norell}.  Guarded recursion \cite{Nakano}
is a different approach where the type system is enhanced with a
modality, called ``later'' and written \IC{▻}, encoding time delay in
types. The later modality comes with a general fixpoint combinator for
programming with productive recursive functions and allows the
specification of guarded recursive types, \eg the type of guarded
streams satisfying the isomorphism \F{Str} \Ar{A} \F{≃} \Ar{A} \F{×}
\IC{▻} (\F{Str} \Ar{A}). The tail of a guarded stream is available only
one time step ahead from now, as evidenced by the application of the
later modality to \F{Str} \Ar{A}. Coinductive types can be constructed
from guarded recursive types in type theories that additionally
support clock variables and universal quantification over them
\cite{atkey2013productive,BahrGM17}. Currently there is no
implementation of a calculus with guarded recursion.
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory with guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aformentioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant''.  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. In this respect \GTT\ is
closer in spirit to Atkey and McBride's language. The design of \GTT\
also has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

@@

This paper only include the essential parts of our Agda
formalization. The full code is available at
\url{https://github.com/niccoloveltri/agda-clott}. The formalization
uses Agda 2.5.4.1 and Agda standard library 0.16. The paper is
extracted from a literate Agda file, which implies that all the
displayed code has passed Agda's type and termination checker.

### 2019-01-29 code-oneclock-paper/latex/paper.tex
Some rewriting (Sections 1-4)

\newcommand{\GTT}{\textbf{GTT}}
\newcommand{\Set}{\mathbf{Set}}

@@

\subsection{Modelling Simple Types}
%\label{sec:kripke}
\input{CloTT/Structure/Contexts}
\input{CloTT/Structure/Types}
\input{CloTT/Structure/Terms}
\input{CloTT/Structure/Subst}
\input{CloTT/Structure/DefinitionalEquality}
\input{CloTT/Structure/ContextOperations}

Lastly, we use operations on sets and presheaves to define the type operators in simple type theory.
For sets, we use products, sums, and function types in Agda.
Products, sums, and exponentials of presheaves are interpreted with the operations in \Cref{sec:presheaves}.
We only give the notation for function types.

\input{CloTT/TypeFormers/FunctionType}

>>>>

\newcommand{\GTT}{\textbf{GTT}}
\newcommand{\AM}{\textbf{AM}}
\newcommand{\Set}{\mathbf{Set}}

@@

\subsection{Modelling Simple Types}
%\label{sec:kripke}
\input{CloTT/Structure/Contexts}
\input{CloTT/Structure/Types}
\input{CloTT/Structure/Terms}
\input{CloTT/Structure/Subst}
\input{CloTT/Structure/DefinitionalEquality}

To interpret simple type theory, we also need to give the context operations, simple type formers, and operations on substitutions.
Since their definitions are standard, we do not discuss them formally.
For each of them, we need to make a case distinction based on the clock context.
For example, the empty context is interpreted by the unit type and terminal presheaf depending on the clock context.
We use the operations on presheaves defined in \Cref{sec:presheaves} to interpret simple type formers.
\input{CloTT/TypeFormers/FunctionType}
